---
title: "Football report"
output: html_document
name: Your name here
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Report overview
This report summarises player data for football matches contested in 2014 and 2015. In total there are 368 observations in the dataset.

### Sprint distance by playing position
First, clean up the data by changing any text variables (e.g., home/away) into factor/categorical ones. The "phase_of_season" variable is coded as a number (1-4). Change this to a factor too.

Then, create a boxplot depicting the sprint distance (sprint_distance) covered during matches by each playing position (position). Which playing position tends to cover the highest amount of sprint distance during matches? Hint: set the fill to "position" to produce different coloured boxplots for each playing position.

```{r distance-by-position, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(ggplot2)

football_data <- read_csv(file = "H:/ads-v2/data/formative_data.csv")

football_data <- football_data %>% 
  # clean up the variable names to make them easier to work with
  clean_names() %>% 
  mutate_if(is.character, as.factor) %>% 
  mutate(phase_of_season = as.factor(phase_of_season))

ggplot(football_data) +
  aes(x = position, y = sprint_distance_m, fill = position) +
  geom_boxplot() +
  scale_fill_hue(direction = 1) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ylab("Sprint distance (m)") +
  xlab("Playing position")
```

### High speed running over the season by playing position
Then, create a figure that shows how the mean amount of High Speed Running (high_speed_distance_5_5_7m_s) covered by each playing position changed over the phases of the season (phase_of_season). Does it look as though any of the playing positions were getting more fatigued as the season went on? Hint: use the facet_wrap() function to produce a separate plot for each level of the "position" variable.

```{r running-over-season-by-position, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(football_data) +
 aes(x = phase_of_season, y = high_speed_distance_5_5_7m_s, fill = position) +
 geom_bar(stat="summary", fun="mean") +
 scale_fill_hue(direction = 1) +
 theme_minimal() +
 facet_wrap(vars(position)) +
 ylab("Mean high speed running distance (m)") +
 xlab("Phase of season") +
 theme(panel.spacing.y = unit(1, "lines"))
```

### Match outcomes for home and away games
Generate a two-way frequency table (or cross tabulation) of match outcomes (i.e., draw, loss, win) by match location (i.e., home or away). Edit your table using the <pkg>kableExtra</pkg> package. Is the total number of losses greater at home or away games?

```{r outcomes-by-location, echo=FALSE, message=FALSE, warning=FALSE}
library(kableExtra)
wins_location <- table(football_data$match_outcome, football_data$match_location)
kable <- kable(wins_location)
k <- kable(wins_location, 
      caption = "Table 1. Match outcomes for home and away games.", 
      align = "c")
k_style <- kable_classic(k, full_width = TRUE) 
k_highlighted <- row_spec(k_style, row = 0, bold = TRUE, color = "white",background="#FF5733", underline=TRUE) 
k_highlighted
```
